Some examples in extended logic counting classes

### 2016-12-29 LICS/appendix.tex
29 dic

If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$. We also denote $\length(\bar{x}) = m$ ($\length(\bar{X}) = m$).

\subsection{Proof of Theorem \ref{no-mult}}

@@

\subsection{Proof of Theorem \ref{captfp}}

@@

\subsection{Proof of Theorem \ref{qfo-pfp-cap}}

@@

\subsection{Proof of Theorem \ref{eqso-fo-cap}}

@@

\subsection{Proof of Theorem \ref{eqso-eso-cap}}

@@

\subsection{Proof of Theorem \ref{qso-pfp-cap}}

@@

\subsection{Proof of Theorem \ref{eqso-sigma-zero-in-fp}}

@@

\subsection{Proof of Theorem \ref{eqso-sigma-one-in-eqso-pi-one}}

@@

%\subsection{Proof of Theorem \ref{one-sigma-zero}}
%
%Towards a contradiction, suppose that there exists a constant function $f(\A) = s > 0$ which is expressible in $\E{0}$. This is, there is a formula $\psi(x,X)$ such that $f_{\psi(x,X)}(\A) = \vert \{ \langle e,P\rangle \mid \A\models\psi(e,P) \} \vert = s$ for each $\A\in\ostr$. It can be seen that if $X$ contains more than zero variables, then the function $f_{\psi(x,X)}$ grows exponentially. We have now that $\psi(x)$ contains only first order variables. Consider some $\A \in \ostr$ of size $n$ and let $a_1,\ldots,a_s\in \A$ be the tuples that satisfy $\A \models \psi(a_i)$. Now consider a structure $\A'$ that is obtained by duplicating $\A$. This is, $\A'$ has size $2n$, its first half is isomorphic to its second half, and also to $\A$. We have that there is at least $2s$ tuples $a_i,b_i\in \A'$, $s$ on each half, such that $\A'\models\psi(a_i)$ and $\A'\models\psi(b_i)$, which leads to a contradiction.

%\textbf{Proof of Theorem \ref{mult-sigma-zero}}
%
%Let $\alpha \in \sqso(\loge{0})$ with binary product. This is, $\alpha$ is defined by the grammar $\alpha := \sa{x}\varphi \ \mid \ \sa{X}\varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha \ \mid \ (\alpha\mult\alpha)$, where $\varphi \in \loge{0}$. For convenience, we add the expressions $\sa{x}(\varphi\mult\varphi)$ and $\sa{X}(\varphi\mult\varphi)$ to the grammar. Let $k$ be the number of algebraic operators in $\alpha$. We will show by induction on $k$ that $\alpha$ is expressible without multiplication. If $k = 1$, then the possible cases are $\alpha = \sa{x}\varphi$, where $\varphi\in\loge{0}$, which is already product-free. If $k = 2$, then the only that is not two nested summations is $\alpha = \sa{x}(\varphi\mult\psi)$, which can be expressed as $\alpha = \sa{x}(\varphi\wedge\psi)$. Suppose that the hypothesis holds for $k = 1$ through $n-1$. Let $\alpha$ have $n$ algebraic operators. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$ where $\beta$ is an algebraic expression, then $\beta$ has $n-1$ algebraic operators and by hypothesis it can be expressed as product-free. If $\alpha = (\beta \mult \gamma)$ there are two cases. If at least one between $\beta$ and $\gamma$ has a product, then by hypothesis it can be expressed as product-free. Moreover, the new expression for $\alpha$ has strictly less than $n$ algebraic operators so by hypothesis it can be expressed as product-free. If there are no products in $\beta$ or $\gamma$, then let $\beta = \sa{x}\delta(x)$. We have that $\alpha$ can be expressed as $\sa{y}(\delta(y)\mult\gamma)$, where $y$ is a first-order variable that does not appear in $\beta$ or $\gamma$. Since $\delta(y)\mult\gamma$ has $n-1$ algebraic operators, it can be expressed as product-free, and then $\alpha$ can be expressed as product-free.
%
%\textbf{Proof of Theorem \ref{fo-prod-sigma-zero}}
%
%The resulting grammar is $\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha \ \mid \pa{x}\alpha$. Consider a more restricted grammar that replaces $\pa{x}\alpha$ by $\pa{x}\sa{y}\varphi$. Let $\alpha = \pa{x}\sa{y}(x = y)$. This formula produces a constant function equal to 1, which as shown in Theorem \ref{one-sigma-zero} is not expressible in $\E{0}$. 
%
%\textbf{Proof of Theorem \ref{fp1}}
%
%Let $f \in \XE{0}$, and let $\varphi(\X,\x)$ be and {\sc FO}-extended quantifier-free $\L$-formula such that:
%\begin{eqnarray*}
%	f(\A) &=& \vert \{\langle \P,\e  \rangle \mid \A \models \varphi(\P,\e) \} \vert
%\end{eqnarray*}
%for each $\A = \langle A, \S^{\A}, \leq^{\A} \rangle \in \str$, where $\e \in A^m$ and $\P = (P_1,\ldots,P_q)$ is a tuple of predicates. We will now show that computing $f(\A)$ can be done in polynomial time.
%
%For each {\sc FO}-formula $\beta(\x)$ in $\varphi(\X,\x)$, let $R_{\beta}$ be a predicate of arity $m$. Let $\A^\prime = \langle A, \S^{\A}, R_{\beta}^{\A^\prime}, \leq^{\A} \rangle \in \str$, where $R_{\beta}^{\A^\prime} = \{\d \mid \A \models \beta(\d)\}$. Note that each $\beta(\x)$ is fixed in $\varphi(\X,\x)$, and for each $\d \in A^m$, checking whether $\A \models \beta(\d)$ can be done in polynomial time. Therefore, generating $R_{\beta}^{\A^\prime}$ can also be done in polynomial time.
%
%Let $\psi(\X,\x)$ be obtained by replacing each {\sc FO}-formula $\beta(\x)$ in $\varphi(\X,\x)$ by $R_{\beta}(\x)$. Also, let $g = f_{\psi(\X,\x)}$. Note that for each tuple of predicates $\P$ and each $\e \in A^m$, $\A \models \varphi(\P,\e)$ if and only if $\A^\prime \models \psi(\P,\e)$, and so, $g(\A^\prime) = f(\A)$. But, $\psi(\X,\x)$ is a quantifier-free $\L$-formula, and therefore, $g \in \#\Sigma_0$. Since it is shown in \cite{DBLP:journals/jcss/SalujaST95} that $\#\Sigma_0 \subseteq$ {\sc FP}, we conclude that $f(\A)$ can be evaluated in polynomial time.

\textbf{Proof of Theorem \ref{decisionptime}}

@@

\textbf{Proof of Theorem \ref{sub-pnp}}

@@

%\textbf{Proof of Theorem \ref{sigma1strict}}
%
%We will show that the $\E{1}$ function $f$ defined by $\varphi(x_1) = (x_1 = x_1) \wedge \forall y \, S(y)$ is not in $\#\Sigma_1$. By contradiction, suppose that it is. Let $\A = \langle A =\{1\},S^\A = \{1\},\leq^\A=\{(1,1)\} \rangle$. Then, $f(\A) = 1$. Now let $\B = \langle B =\{1,2\},S^\B = \{1\},\leq^\B=\{(1,1),(1,2),(2,2)\} \rangle$. Note that $\A$ is an induced substructure of $\B$.
%
%We have that for each function $g \in \#\Sigma_1$ and structures $\A_1,\A_2\in\str$, if $\A_1$ is an induced substructure of $\A_2$, then $g(\A_1) \leq g(\A_2)$ \cite{SalujaST95}. Therefore, $f(\B) \geq f(\A) = 1$. However, there is no assignment $s\in B$ to $x$ such that $\B\models\:\varphi(s)$, so $f(\B) = 0$, which leads to a contradiction.

\subsection{Proof of Theorem \ref{sigmafo-minusone}}

@@

%\textbf{Proof of Theorem \ref{dnf-pars}}

@@

%\textbf{Proof of Theorem \ref{nplusone-strict}}

@@

%\textbf{Proof of Theorem \ref{sigma2-pe}}

@@

\textbf{Proof of Theorem \ref{sigma2hard}}

@@

Consider a finite $\R$-structure $\A$ with domain $A$. To simplify the proof, we extend our grammar to allow first-order constants. Consider each tuple $\bar{a}\in A^{\length(\bar{x})}$, each $\bar{b}\in A^{\length(\bar{y})}$ and each $\bar{c}\in A^{\length(\bar{z})}$ as a tuple of first-order constants. The following formula defines the same function as $\alpha$:
$$
c + \sum_{i = 1}^{\#clauses} \sum_{\bar{a}\in A^{\length(\bar{x})}} \sa{\bar{X}}\bigvee_{\bar{b}\in A^{\length(\bar{y})}}\bigwedge_{j = 1}^{n}\bigwedge_{\bar{c}\in A^{\length(\bar{z})}}\varphi^i_j(\bar{X},\bar{a},\bar{b},\bar{c}).
$$
Note that each $\fo$ formula in each $\varphi^i_j$ over $(\bar{x},\bar{y},\bar{z})$ has no free variables. Therefore, we can evaluate each of these in polynomial time and replace them by $\perp$ and $\top$ where it corresponds. Each $\varphi^i_j$ will be of the form $\perp \vee\, \chi^i_j(\bar{X})$ or $\top \vee \chi^i_j(\bar{X})$ where $\chi^i_j$ is a disjunction of $\neg X_{\ell}$'s and at most one $X_{\ell}$. The formulas of the form $\top \vee \chi^i_j(\bar{X})$ can be removed entirely, and the formulas of the form $\perp \vee\, \chi^i_j(\bar{X})$ can be replaced by $\chi^i_j(\bar{X})$. We obtain the formula
$$
c + \sum_{i = 1}^{m}\sa{\bar{X}}\bigvee_{j = 1}^{\#d}\bigwedge_{k = 1}^{\#c}\psi^{i}_{j,k}(\bar{X})
$$
where every $\psi^{i}_{j,k}(\bar{X})$ is a disjunction of $\neg X_{\ell}$'s and zero or one $X_{\ell}$.

@@

If $\alpha$ is equal to a constant $c$, then we define $g_{\alpha}(\A)$ as the following formula that has exactly $c$ satisfying assignments:
$$
g_{\alpha}(\A) = \bigvee_{i = 1}^{c}\neg t_1 \wedge \cdots \wedge \neg t_{i-1} \wedge t_i \wedge \neg t_{i+1} \wedge \cdots \wedge \neg p_c.
$$ 
If $\alpha = \beta_1 + \cdots + \beta_m$, let $g_{\beta_i}(\A) = \bigvee_{j = 1}^{\#d}\bigwedge_{k = 1}^{\#c}\theta^i_{i,j}$ for each $\beta_i$ where each $\theta^i_{j,k}$ is a Horn clause. Let $\Theta_i = g_{\beta_i}(\A)$. We rename the variables in each $\Theta_i$ so none of them are mentioned more than once in total. We add $m$ new variables $t_1,\ldots,t_m$ and we define:
\begin{align*}
g_{\alpha}(\A) = &\bigvee_{i = 1}^{\#d}(\bigwedge_{j = 1}^{\#c}\theta^1_{i,j} \wedge (\bigwedge\limits_{\substack{\text{each } t\\ \text{ mentioned in}\\ \Theta_2,\ldots,\Theta_{m}}}t) \wedge (t_1 \wedge \bigwedge_{\ell = 2}^{m} \neg t_{\ell})) \vee \\ 
&\bigvee_{i = 1}^{\#d}(\bigwedge_{j = 1}^{\#c}\theta^2_{i,j} \wedge (\bigwedge\limits_{\substack{\text{each $t$}\\ \text{ mentioned in}\\ \Theta_1,\Theta_3,\ldots,\Theta_{m}}}t) \wedge (t_2 \wedge \bigwedge\limits_{\substack{\ell = 1 \\ \ell \neq 2}}^{m} \neg t_{\ell})) \vee \cdots \vee\\ 
&\bigvee_{i = 1}^{\#d}(\bigwedge_{j = 1}^{\#c}\theta^m_{i,j} \wedge (\bigwedge\limits_{\substack{\text{each } t\\ \text{ mentioned in}\\ \Theta_2,\ldots,\Theta_{m-1}}}t) \wedge (t_m \wedge \bigwedge_{\ell = 1}^{m-1} \neg t_{\ell})).
\end{align*}
The number of satisfying assignments for this formula is exactly the sum of satisfying assignments for all $g_{\beta_i}(\A)$. This covers all possible cases for $\alpha$, and the complete procedure takes polynomial time.

@@

\textbf{Proof of Theorem \ref{so-rec}}

\textbf{Proof of Theorem \ref{tqfo-fo-cap}}

@@

\textbf{Proof of Theorem \ref{tqso-fo-cap}}

\textbf{Proof of Theorem \ref{rqfo-fo-cap}}

>>>>

If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$. We also denote $\length(\bar{x}) = m$ ($\length(\bar{X}) = m$).
\subsection*{Proof of Theorem \ref{no-mult}}

@@

\subsection*{Proof of Theorem \ref{captfp}}

@@

\subsection*{Proof of Theorem \ref{qfo-pfp-cap}}

@@

\subsection*{Proof of Theorem \ref{eqso-fo-cap}}

@@

\subsection*{Proof of Theorem \ref{eqso-eso-cap}}

@@

\subsection*{Proof of Theorem \ref{qso-pfp-cap}}

@@

\subsection*{Proof of Theorem \ref{eqso-sigma-zero-in-fp}}

@@

\subsection*{Proof of Theorem \ref{eqso-sigma-one-in-eqso-pi-one}}

@@

%\subsection*{Proof of Theorem \ref{one-sigma-zero}}
%
%Towards a contradiction, suppose that there exists a constant function $f(\A) = s > 0$ which is expressible in $\E{0}$. This is, there is a formula $\psi(x,X)$ such that $f_{\psi(x,X)}(\A) = \vert \{ \langle e,P\rangle \mid \A\models\psi(e,P) \} \vert = s$ for each $\A\in\ostr$. It can be seen that if $X$ contains more than zero variables, then the function $f_{\psi(x,X)}$ grows exponentially. We have now that $\psi(x)$ contains only first order variables. Consider some $\A \in \ostr$ of size $n$ and let $a_1,\ldots,a_s\in \A$ be the tuples that satisfy $\A \models \psi(a_i)$. Now consider a structure $\A'$ that is obtained by duplicating $\A$. This is, $\A'$ has size $2n$, its first half is isomorphic to its second half, and also to $\A$. We have that there is at least $2s$ tuples $a_i,b_i\in \A'$, $s$ on each half, such that $\A'\models\psi(a_i)$ and $\A'\models\psi(b_i)$, which leads to a contradiction.

%\subsection*{Proof of Theorem \ref{mult-sigma-zero}}
%
%Let $\alpha \in \sqso(\loge{0})$ with binary product. This is, $\alpha$ is defined by the grammar $\alpha := \sa{x}\varphi \ \mid \ \sa{X}\varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha \ \mid \ (\alpha\mult\alpha)$, where $\varphi \in \loge{0}$. For convenience, we add the expressions $\sa{x}(\varphi\mult\varphi)$ and $\sa{X}(\varphi\mult\varphi)$ to the grammar. Let $k$ be the number of algebraic operators in $\alpha$. We will show by induction on $k$ that $\alpha$ is expressible without multiplication. If $k = 1$, then the possible cases are $\alpha = \sa{x}\varphi$, where $\varphi\in\loge{0}$, which is already product-free. If $k = 2$, then the only that is not two nested summations is $\alpha = \sa{x}(\varphi\mult\psi)$, which can be expressed as $\alpha = \sa{x}(\varphi\wedge\psi)$. Suppose that the hypothesis holds for $k = 1$ through $n-1$. Let $\alpha$ have $n$ algebraic operators. If $\alpha = \sa{x}\beta$ or $\alpha = \sa{X}\beta$ where $\beta$ is an algebraic expression, then $\beta$ has $n-1$ algebraic operators and by hypothesis it can be expressed as product-free. If $\alpha = (\beta \mult \gamma)$ there are two cases. If at least one between $\beta$ and $\gamma$ has a product, then by hypothesis it can be expressed as product-free. Moreover, the new expression for $\alpha$ has strictly less than $n$ algebraic operators so by hypothesis it can be expressed as product-free. If there are no products in $\beta$ or $\gamma$, then let $\beta = \sa{x}\delta(x)$. We have that $\alpha$ can be expressed as $\sa{y}(\delta(y)\mult\gamma)$, where $y$ is a first-order variable that does not appear in $\beta$ or $\gamma$. Since $\delta(y)\mult\gamma$ has $n-1$ algebraic operators, it can be expressed as product-free, and then $\alpha$ can be expressed as product-free.
%
%\subsection*{Proof of Theorem \ref{fo-prod-sigma-zero}}
%
%The resulting grammar is $\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha \ \mid \pa{x}\alpha$. Consider a more restricted grammar that replaces $\pa{x}\alpha$ by $\pa{x}\sa{y}\varphi$. Let $\alpha = \pa{x}\sa{y}(x = y)$. This formula produces a constant function equal to 1, which as shown in Theorem \ref{one-sigma-zero} is not expressible in $\E{0}$. 
%
%\subsection*{Proof of Theorem \ref{fp1}}
%
%Let $f \in \XE{0}$, and let $\varphi(\X,\x)$ be and {\sc FO}-extended quantifier-free $\L$-formula such that:
%\begin{eqnarray*}
%	f(\A) &=& \vert \{\langle \P,\e  \rangle \mid \A \models \varphi(\P,\e) \} \vert
%\end{eqnarray*}
%for each $\A = \langle A, \S^{\A}, \leq^{\A} \rangle \in \str$, where $\e \in A^m$ and $\P = (P_1,\ldots,P_q)$ is a tuple of predicates. We will now show that computing $f(\A)$ can be done in polynomial time.
%
%For each {\sc FO}-formula $\beta(\x)$ in $\varphi(\X,\x)$, let $R_{\beta}$ be a predicate of arity $m$. Let $\A^\prime = \langle A, \S^{\A}, R_{\beta}^{\A^\prime}, \leq^{\A} \rangle \in \str$, where $R_{\beta}^{\A^\prime} = \{\d \mid \A \models \beta(\d)\}$. Note that each $\beta(\x)$ is fixed in $\varphi(\X,\x)$, and for each $\d \in A^m$, checking whether $\A \models \beta(\d)$ can be done in polynomial time. Therefore, generating $R_{\beta}^{\A^\prime}$ can also be done in polynomial time.
%
%Let $\psi(\X,\x)$ be obtained by replacing each {\sc FO}-formula $\beta(\x)$ in $\varphi(\X,\x)$ by $R_{\beta}(\x)$. Also, let $g = f_{\psi(\X,\x)}$. Note that for each tuple of predicates $\P$ and each $\e \in A^m$, $\A \models \varphi(\P,\e)$ if and only if $\A^\prime \models \psi(\P,\e)$, and so, $g(\A^\prime) = f(\A)$. But, $\psi(\X,\x)$ is a quantifier-free $\L$-formula, and therefore, $g \in \#\Sigma_0$. Since it is shown in \cite{DBLP:journals/jcss/SalujaST95} that $\#\Sigma_0 \subseteq$ {\sc FP}, we conclude that $f(\A)$ can be evaluated in polynomial time.

\subsection*{Proof of Theorem \ref{decisionptime}}

@@

\subsection*{Proof of Theorem \ref{sub-pnp}}

@@

%\subsection*{Proof of Theorem \ref{sigma1strict}}
%
%We will show that the $\E{1}$ function $f$ defined by $\varphi(x_1) = (x_1 = x_1) \wedge \forall y \, S(y)$ is not in $\#\Sigma_1$. By contradiction, suppose that it is. Let $\A = \langle A =\{1\},S^\A = \{1\},\leq^\A=\{(1,1)\} \rangle$. Then, $f(\A) = 1$. Now let $\B = \langle B =\{1,2\},S^\B = \{1\},\leq^\B=\{(1,1),(1,2),(2,2)\} \rangle$. Note that $\A$ is an induced substructure of $\B$.
%
%We have that for each function $g \in \#\Sigma_1$ and structures $\A_1,\A_2\in\str$, if $\A_1$ is an induced substructure of $\A_2$, then $g(\A_1) \leq g(\A_2)$ \cite{SalujaST95}. Therefore, $f(\B) \geq f(\A) = 1$. However, there is no assignment $s\in B$ to $x$ such that $\B\models\:\varphi(s)$, so $f(\B) = 0$, which leads to a contradiction.

\subsection*{Proof of Theorem \ref{sigmafo-minusone}}

@@

%\subsection*{Proof of Theorem \ref{dnf-pars}}

@@

%\subsection*{Proof of Theorem \ref{nplusone-strict}}

@@

%\subsection*{Proof of Theorem \ref{sigma2-pe}}

@@

\subsection*{Proof of Theorem \ref{sigma2hard}}

@@

Consider a finite $\R$-structure $\A$ with domain $A$. To simplify the proof, we extend our grammar to allow first-order constants. Consider each tuple $\bar{a}\in A^{\length(\bar{x})}$, each $\bar{b}\in A^{\length(\bar{y})}$ and each $\bar{c}\in A^{\length(\bar{z})}$ as a tuple of first-order constants. The following formula defines the same function as $\alpha$:
$$
c + \sum_{i = 1}^{\#clauses} \sum_{\bar{a}\in A^{\length(\bar{x})}} \sa{\bar{X}}\bigvee_{\bar{b}\in A^{\length(\bar{y})}}\bigwedge_{j = 1}^{n}\bigwedge_{\bar{c}\in A^{\length(\bar{z})}}\varphi^i_j(\bar{X},\bar{a},\bar{b},\bar{c}).
$$
Note that each $\fo$ formula over $(\bar{x},\bar{y},\bar{z})$ in each $\varphi^i_j$ has no free variables. Therefore, we can evaluate each of these in polynomial time and replace them by $\perp$ and $\top$ where it corresponds. Each $\varphi^i_j$ will be of the form $\perp \vee\, \chi^i_j(\bar{X})$ or $\top \vee \chi^i_j(\bar{X})$ where $\chi^i_j$ is a disjunction of $\neg X_{\ell}$'s and at most one $X_{\ell}$. The formulas of the form $\top \vee \chi^i_j(\bar{X})$ can be removed entirely, and the formulas of the form $\perp \vee\, \chi^i_j(\bar{X})$ can be replaced by $\chi^i_j(\bar{X})$. We obtain the formula
$$
c + \sum_{i = 1}^{m}\sa{\bar{X}}\bigvee_{j = 1}^{\#d}\bigwedge_{k = 1}^{\#c}\psi^{i}_{j,k}(\bar{X})
$$
where every $\psi^{i}_{j,k}(\bar{X})$ is a disjunction of $\neg X_{\ell}$'s and zero or one $X_{\ell}$.

@@

If $\alpha$ is equal to a constant $c$, then we define $g_{\alpha}(\A)$ as the following formula that has exactly $c$ satisfying assignments:
$$
g_{\alpha}(\A) = \bigvee_{i = 1}^{c}\neg t_1 \wedge \cdots \wedge \neg t_{i-1} \wedge t_i \wedge \neg t_{i+1} \wedge \cdots \wedge \neg p_c.
$$ 
If $\alpha = \beta_1 + \cdots + \beta_m$, let $g_{\beta_i}(\A) = \bigvee_{j = 1}^{\#d}\bigwedge_{k = 1}^{\#c}\theta^i_{i,j}$ for each $\beta_i$ where each $\theta^i_{j,k}$ is a Horn clause. Let $\Theta_i = g_{\beta_i}(\A)$. We rename the variables in each $\Theta_i$ so none of them are mentioned in any other $\Theta_j$. We add $m$ new variables $t_1,\ldots,t_m$ and we define:
\begin{align*}
g_{\alpha}(\A) = &\bigvee_{i = 1}^{\#d}(\bigwedge_{j = 1}^{\#c}\theta^1_{i,j} \wedge (\bigwedge\limits_{\substack{\text{each } t\\ \text{ mentioned in}\\ \Theta_2,\ldots,\Theta_{m}}}t) \wedge (t_1 \wedge \bigwedge_{\ell = 2}^{m} \neg t_{\ell})) \vee \\ 
&\bigvee_{i = 1}^{\#d}(\bigwedge_{j = 1}^{\#c}\theta^2_{i,j} \wedge (\bigwedge\limits_{\substack{\text{each $t$}\\ \text{ mentioned in}\\ \Theta_1,\Theta_3,\ldots,\Theta_{m}}}t) \wedge (t_2 \wedge \bigwedge\limits_{\substack{\ell = 1 \\ \ell \neq 2}}^{m} \neg t_{\ell})) \vee \cdots \vee\\ 
&\bigvee_{i = 1}^{\#d}(\bigwedge_{j = 1}^{\#c}\theta^m_{i,j} \wedge (\bigwedge\limits_{\substack{\text{each } t\\ \text{ mentioned in}\\ \Theta_2,\ldots,\Theta_{m-1}}}t) \wedge (t_m \wedge \bigwedge_{\ell = 1}^{m-1} \neg t_{\ell})).
\end{align*}
The formula is a disjunction of Horn formulas, and the number of satisfying assignments for this formula is exactly the sum of satisfying assignments for each $g_{\beta_i}(\A)$. This, at the same time, is equal to $\sem{\alpha}(\A)$. This covers all possible cases for $\alpha$, and the entire procedure takes polynomial time.

@@

\subsection*{Proof of Theorem \ref{so-rec}}

\subsection*{Proof of Theorem \ref{tqfo-fo-cap}}

@@

\subsection*{Proof of Theorem \ref{tqso-fo-cap}}

\subsection*{Proof of Theorem \ref{rqfo-fo-cap}}
